---
title: "Introduction to transmission suitability maps dataset"
header-includes:
    - \usepackage{setspace}
fontsize: 10pt
output:
  pdf_document:
  html_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

## Preambule

In this document, we provide some example code for the handling and visualization of the transmission suitability maps for DENV transmitted by *Aedes aegypti* mosquitoes. These scripts are meant to provide a brief introduction to the files associated with the dataset and some potential methods of managing the data. 

## Set up

First, we load useful packages. 

\small
```{r message=FALSE}
library(raster)
library(tidyverse)
library(RColorBrewer)
```
\normalsize

We also define the file path to the local directory where the dataset of maps are stored. 

\small
```{r}
### INSERT FILE PATH HERE
indexP_maps_path <- ""
```
\normalsize

## Typical year maps

The typical year maps provide an estimate of the seasonal dynamics of Index P for a "typical" calendar year from 1981 to 2019. Here, we display the seasonality of DENV transmission potential during a typical year across Brazil.

\small
```{r fig.height=5, fig.width=10}
# `plot_typical_year_maps` takes as an argument the string for the selected country/territory 
# and plots a grid of 12 Index P maps, one for each calendar month. 
plot_typical_year_maps <- function(country) {
  typical_year_map_ext <- "_indexP_typical_year_mean_rasters.tif"
  file_name <- paste0(country, typical_year_map_ext)
  file_path <- file.path(indexP_maps_path, country, file_name)
  
  mth_mapping <- month.abb
  names(mth_mapping) <- paste0("X", 1:12)
  indexP_typical_year_mean_rasters <- raster::brick(file_path) %>% 
    rasterToPoints() %>% 
    as.data.frame() %>% 
    gather(key="layer", value="indexP", "X1":"X12") %>% 
    mutate(layer=factor(mth_mapping[layer], levels=month.abb))

  # generate outline of Brazil
  col_scale <- c(colorRampPalette(rev(brewer.pal(11, "RdYlBu"))[1:5])(5), 
                 colorRampPalette(rev(brewer.pal(11, "RdYlBu"))[6:11])(15), 
                 rep(rev(brewer.pal(11, "RdYlBu"))[11], 5))
  world_map <- map_data("world", regions=country)
  g <- ggplot(data=world_map) + 
    geom_map(map=world_map, mapping=aes(map_id=country), 
             color="black", fill="lightgrey", size=1) + 
    expand_limits(x = world_map$long, y = world_map$lat) + 
    geom_tile(data=indexP_typical_year_mean_rasters, 
              mapping=aes(x=x, y=y, fill=indexP)) + 
    scale_fill_gradientn(colours=rep(col_scale, each=3), 
                         na.value="white", 
                         guide=guide_colorbar(frame.colour = "black", ticks.colour = "black"), 
                         limits=c(0, 4), breaks=seq(0, 4, 1), name="Index P") + 
    theme(panel.border=element_blank(), panel.grid=element_blank(), 
          axis.title=element_blank(), axis.text=element_blank(), 
          axis.ticks=element_blank(), panel.background=element_blank(), 
          strip.text=element_text(size=9, family="sans"), 
          legend.title=element_text(size=9, family="sans"), 
          legend.text=element_text(size=9, family="sans")) + 
    facet_wrap(~layer, nrow=2)
  return(g)
}
## maps for Brazil
plot_typical_year_maps(country="Brazil")
```
\normalsize

## Monthly time series

We might also be interested in the dynamics of Index P over many years for a particular region. Here, we show the monthly DENV transmission potential in Rio de Janeiro, Brazil from 1981 to 2019. 

\small
```{r fig.height=3, fig.width=6}
# `plot_pixel_time_series` takes as arguments a country as well as longitude and latitude 
# coordinates within the bounds of that country. It returns a plot of the monthly 
# DENV transmission potential from 1981 to 2019 for the closest spatial pixel. 
plot_pixel_time_series <- function(country, lon, lat) {
  monthly_time_series_ext <- "_indexP_monthly_mean_rasters.tif"
  file_name <- paste0(country, monthly_time_series_ext)
  file_path <- file.path(indexP_maps_path, country, file_name)
  monthly_mean_rasters <- raster::brick(file_path)
  
  # retrieve data for closest spatial pixel
  closest_pixel <- monthly_mean_rasters[[1]] %>% 
    rasterToPoints(., spatial = TRUE) %>% 
    as.data.frame() %>% 
    mutate(dist=(lat-y)^2 + (lon-x)^2) %>% 
    arrange(-desc(dist)) %>% 
    head(1)
  location <- data.frame(x=closest_pixel$x, y=closest_pixel$y)
  coordinates(location) <- ~x+y
  point = SpatialPoints(location,proj4string = CRS("+init=epsg:4326"))
  time_series = raster::extract(monthly_mean_rasters, point) %>% as.numeric()
  dates <- sub('.', '', names(monthly_mean_rasters)) %>% as.Date(., format=c("%Y.%m.%d"))
  time_series <- data.frame(date=dates, indexP=time_series)
  
  g <- ggplot(data=time_series) + 
    geom_hline(yintercept=1, color="black", linetype="dashed") + 
    geom_line(mapping=aes(x=date, y=indexP)) + 
    scale_x_date(breaks=seq(as.Date("1980/1/1"), as.Date("2020/1/1"), "5 years"), 
                 date_labels = "'%y", limits=as.Date(c("1980-01-01", "2020-03-01")), 
                 expand=c(0.01, 0.7)) + 
    labs(y="Mean index P") + 
    theme_minimal() + 
    theme(axis.title.x=element_blank(), 
          axis.line=element_line(),
          axis.text=element_text(size=8, family="sans", color="black"),
          panel.border=element_blank(), 
          panel.grid=element_blank(),
          axis.title=element_text(size=9, family="sans", color="black"), 
          axis.ticks=element_line(color="black"), panel.spacing.y=unit(-0.2, "cm"))
  return(g)
}
## monthly time series for Rio de Janerio, Brazil
plot_pixel_time_series(country="Brazil", lon=-43.196388, lat=-22.908333)
```
\normalsize
